---
title: "Mapping ABS Database to Darwin Core"
author: "Antigravity (Pair Programming with Dimitri Brosens)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(odbc)
library(DBI)
library(sf)
library(leaflet)
library(lubridate)
library(janitor)
```

## Introduction

This report documents the workflow for mapping the **Autochtone Boomsoorten (ABS)** Microsoft Access database to the **Darwin Core (DwC)** standard. The objective is to produce a standardized occurrence dataset for the Vlaams Biodiversiteitsportaal.

## 1. Load Data from Access

We connect to the MS Access database using `odbc`.

```{r connection}
db_path <- "../data/databankABS_v11-03-2013.mdb"

# Connection string for Windows
con <- dbConnect(odbc::odbc(), 
                 .connection_string = paste0("Driver={Microsoft Access Driver (*.mdb, *.accdb)};DBQ=", db_path))

# Load relevant tables
# We use clean_names here to standardize column names to lowercase/snake_case
tblOpnames <- dbReadTable(con, "tblOpnames") %>% clean_names()
tblBomenStruiken <- dbReadTable(con, "tblBomenStruiken") %>% clean_names()
cdeBomenStruiken <- dbReadTable(con, "cdeBomenStruiken") %>% clean_names()

dbDisconnect(con)
```

## 2. Table Joins

In this step, we join the three tables. We take care to resolve any duplicate column names that might arise from the join (like `nummer`).

```{r joins}
# Join tables
# Note: inner_join will add .x and .y suffixes to columns with the same name (except the join keys)
raw_joined_data <- tblOpnames %>%
  inner_join(tblBomenStruiken, by = "proj_cod_loc_nr") %>%
  inner_join(cdeBomenStruiken, by = c("soort" = "soort_code"))

# Debug: Show column names to identify potential issues with suffixes
print("Column names after join:")
print(names(raw_joined_data))
```

## 3. Date Parsing Logic

We define the helper function to parse specifically formatted dates from the `nummer` column.

```{r date-function}
# Helper function for date extraction
extract_full_date <- function(nummer_val, fallback_year) {
  # Early exit for NA or short strings
  if (is.na(nummer_val) || nchar(as.character(nummer_val)) < 6) return(as.character(fallback_year))
  
  prefix <- substr(as.character(nummer_val), 1, 6)
  yr_part <- as.numeric(substr(prefix, 1, 2))
  mo_part <- as.numeric(substr(prefix, 3, 4))
  da_part <- as.numeric(substr(prefix, 5, 6))
  
  # Handle potential NA from numeric conversion
  if (is.na(yr_part) || is.na(mo_part) || is.na(da_part)) return(as.character(fallback_year))
  
  full_yr <- ifelse(yr_part > 50, 1900 + yr_part, 2000 + yr_part)
  
  # Validate date components
  is_valid <- mo_part >= 1 && mo_part <= 12 && 
              da_part >= 1 && da_part <= 31
  
  # Defensive check for comparison with fallback_year
  matches_year <- FALSE
  if (!is.na(fallback_year) && !is.na(full_yr)) {
    matches_year <- (full_yr == as.numeric(fallback_year))
  }
  
  if (is_valid && matches_year) {
    return(sprintf("%04d-%02d-%02d", full_yr, mo_part, da_part))
  }
  return(as.character(fallback_year))
}
```

## 4. Darwin Core Mapping

We apply the mapping logic to split the data into an **Event Core** and an **Occurrence Extension**.

### 4.1 Process Raw Data

First, we prepare the combined dataset with unique IDs and shared fields.

```{r mapping-processing}
# Prepare processed data
processed_data <- raw_joined_data %>%
  mutate(
    # Handle potential suffixes from join
    proj_field = coalesce(
      if ("projectcod" %in% names(.)) projectcod else NULL,
      if ("projectcod.x" %in% names(.)) projectcod.x else NULL,
      if ("project_cod" %in% names(.)) project_cod else NULL
    ),
    nummer_field = coalesce(
      if ("nummer" %in% names(.)) nummer else NULL,
      if ("nummer.x" %in% names(.)) nummer.x else NULL
    ),
    
    # Year extraction
    yr4 = str_extract(proj_field, "\\d{4}"),
    yr2 = str_extract(proj_field, "\\d{2}"),
    extracted_year = case_when(
      !is.na(yr4) ~ yr4,
      !is.na(yr2) ~ ifelse(as.numeric(yr2) > 50, paste0("19", yr2), paste0("20", yr2)),
      TRUE ~ NA_character_
    ),
    
    # Full date extraction
    eventDate = map2_chr(nummer_field, extracted_year, extract_full_date),
    
    # Locality
    locality = paste0(coalesce(gemeente, ""), ": ", coalesce(locatie, "")) %>% str_remove("^: "),
    
    # eventID derived from project/location number
    eventID = paste0("INBO:VBP:ABS:EVENT:", str_remove_all(proj_cod_loc_nr, "\\s+")),
    
    # Status Mapping
    inheems_lower = str_trim(tolower(inheems)),
    establishmentMeans = case_when(
      str_detect(inheems_lower, "[abc]") ~ "native",
      str_detect(inheems_lower, "p") ~ "introduced",
      str_detect(inheems_lower, "s") ~ "uncertain",
      TRUE ~ "uncertain"
    ),
    occurrenceRemarks = case_when(
      str_detect(inheems_lower, "a") ~ "Autochthonous origin",
      str_detect(inheems_lower, "b") ~ "Probably autochthonous origin",
      str_detect(inheems_lower, "c") ~ "Possibly autochthonous origin",
      str_detect(inheems_lower, "p") ~ "Planted / Introduced",
      str_detect(inheems_lower, "s") ~ "Spontaneous origin",
      TRUE ~ "Unknown status"
    ) %>% paste0(" (code: ", inheems, ")")
  ) %>%
  # Global metadata
  mutate(
    basisOfRecord = "HumanObservation",
    kingdom = "Plantae",
    datasetName = "Autochtone boomsoorten en struiken in Vlaanderen",
    institutionCode = "INBO",
    collectionCode = "ABS-monitoring",
    rightsHolder = "Research Institute for Nature and Forest (INBO)",
    license = "http://creativecommons.org/publicdomain/zero/1.0/",
    language = "en",
    samplingProtocol = "Autochtone boomsoorten monitoring"
  )
```

## 5. Coordinate Conversion

We convert Belgian Lambert 72 (EPSG:31370) coordinates to WGS84 (EPSG:4326).

```{r coordinates}
# Filter records with coordinates
valid_coords <- processed_data %>%
  filter(!is.na(coord_hor), !is.na(coord_vert))

# Convert to sf object
abs_sf <- valid_coords %>%
  st_as_sf(coords = c("coord_hor", "coord_vert"), crs = 31370) %>%
  st_transform(4326)

# Extract coordinates and round to 5 decimal places
abs_coords <- st_coordinates(abs_sf) %>%
  as_tibble() %>%
  rename(decimalLongitude = X, decimalLatitude = Y) %>%
  mutate(across(everything(), ~round(., 5)))

# Combine back
processed_final <- valid_coords %>%
  bind_cols(abs_coords)
```

## 6. Create Event Core and Occurrence Extension

Now we split the `processed_final` into two separate dataframes according to the DwC star schema.

```{r splitting}
# 6.1 Event Core
event_core <- processed_final %>%
  select(eventID, eventDate, year = extracted_year, decimalLatitude, decimalLongitude, 
         locality, samplingProtocol, datasetName, institutionCode, collectionCode, 
         rightsHolder, license, language) %>%
  distinct(eventID, .keep_all = TRUE)

# 6.2 Occurrence Extension
occurrence_extension <- processed_final %>%
  mutate(
    # Generate unique occurrenceID within each event
    occurrenceID = paste0("INBO:VBP:ABS:OCC:", str_remove_all(proj_cod_loc_nr, "\\s+"), ":", row_number())
  ) %>%
  select(occurrenceID, eventID, basisOfRecord, scientificName = latijnse_naam, 
         vernacularName = nederlandse_naam, kingdom, individualCount = aantal, 
         establishmentMeans, occurrenceRemarks)

# 6.3 Measurement Or Fact Extension
# We combine site-level (from event) and species-level (from occurrence) measurements

# Site-level measurements (from event core - unique events)
mof_site <- processed_final %>%
  distinct(eventID, .keep_all = TRUE) %>%
  select(eventID, oppervl) %>%
  pivot_longer(cols = oppervl, names_to = "measurementType", values_to = "measurementValue") %>%
  filter(!is.na(measurementValue)) %>%
  mutate(
    measurementType = "surface area",
    measurementUnit = "square meters"
  )

# Species-level measurements (linked to occurrenceID)
mof_species <- processed_final %>%
  # We need to recreate the occurrenceID the same way as in the extension
  mutate(occurrenceID = paste0("INBO:VBP:ABS:OCC:", str_remove_all(proj_cod_loc_nr, "\\s+"), ":", row_number())) %>%
  select(occurrenceID, hoogte_max, omtrek_max, omtrek_sto, verjonging, fl_fr) %>%
  pivot_longer(cols = c(hoogte_max, omtrek_max, omtrek_sto, verjonging, fl_fr), 
               names_to = "measurementType", values_to = "measurementValue") %>%
  filter(!is.na(measurementValue)) %>%
  mutate(
    measurementUnit = case_when(
      measurementType == "hoogte_max" ~ "meters",
      measurementType %in% c("omtrek_max", "omtrek_sto") ~ "centimeters",
      TRUE ~ NA_character_
    ),
    measurementType = case_when(
      measurementType == "hoogte_max" ~ "maximum height",
      measurementType == "omtrek_max" ~ "maximum circumference",
      measurementType == "omtrek_sto" ~ "stool circumference",
      measurementType == "verjonging" ~ "regeneration status",
      measurementType == "fl_fr" ~ "reproductive state",
      TRUE ~ measurementType
    )
  )

# Combine all Move measurements
measurement_or_fact <- bind_rows(
  mof_site %>% mutate(occurrenceID = NA_character_),
  mof_species %>% mutate(eventID = NA_character_)
) %>%
  mutate(measurementID = paste0("INBO:VBP:ABS:MOF:", row_number())) %>%
  select(measurementID, eventID, occurrenceID, measurementType, measurementValue, measurementUnit)
```

## 7. Visualization

Interactive map showing the distribution. For the map, we use the joined data.

```{r map}
leaflet(processed_final %>% head(10000)) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addMarkers(
    lng = ~decimalLongitude, 
    lat = ~decimalLatitude,
    clusterOptions = markerClusterOptions(),
    popup = ~paste0("<b>", nederlandse_naam, "</b><br>", latijnse_naam, "<br>", occurrenceRemarks)
  )
```

## 8. Export

The Darwin Core Event Core and extensions are exported to separate CSV files.

```{r export}
write_csv(event_core, "../data/event_core.csv", na = "")
write_csv(occurrence_extension, "../data/occurrence_extension.csv", na = "")
write_csv(measurement_or_fact, "../data/measurement_or_fact.csv", na = "")
```

---
**Document generated by Antigravity assistant.**
