---
title: "Mapping ABS Database to Darwin Core"
author: "Antigravity (Pair Programming with Dimitri Brosens)"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_float: true
    theme: readable
    highlight: tango
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
library(tidyverse)
library(odbc)
library(DBI)
library(sf)
library(leaflet)
library(lubridate)
library(janitor)
```

## Introduction

This report documents the workflow for mapping the **Autochtone Boomsoorten (ABS)** Microsoft Access database to the **Darwin Core (DwC)** standard. The objective is to produce a standardized occurrence dataset for the Vlaams Biodiversiteitsportaal.

## 1. Load Data from Access

We connect to the MS Access database using `odbc`.

```{r connection}
db_path <- "../data/databankABS_v11-03-2013.mdb"

# Connection string for Windows (ensure Access Driver is installed)
con <- dbConnect(odbc::odbc(), 
                 .connection_string = paste0("Driver={Microsoft Access Driver (*.mdb, *.accdb)};DBQ=", db_path))

# Load relevant tables
tblOpnames <- dbReadTable(con, "tblOpnames") %>% clean_names()
tblBomenStruiken <- dbReadTable(con, "tblBomenStruiken") %>% clean_names()
cdeBomenStruiken <- dbReadTable(con, "cdeBomenStruiken") %>% clean_names()

dbDisconnect(con)
```

## 2. Data Integration and Mapping

We join the tables and apply the Darwin Core mapping logic, including:
- Extracting years from `projectcod`.
- Parsing specific dates from the `nummer` column (YYMMDD format).
- Mapping autochthonous status to `establishmentMeans`.

```{r mapping}
# Helper function for date extraction
extract_full_date <- function(nummer, fallback_year) {
  if (is.na(nummer) || nchar(nummer) < 6) return(as.character(fallback_year))
  
  prefix <- substr(nummer, 1, 6)
  yr_part <- as.numeric(substr(prefix, 1, 2))
  mo_part <- as.numeric(substr(prefix, 3, 4))
  da_part <- as.numeric(substr(prefix, 5, 6))
  
  full_yr <- ifelse(yr_part > 50, 1900 + yr_part, 2000 + yr_part)
  
  # Validate date
  is_valid <- !is.na(mo_part) && mo_part >= 1 && mo_part <= 12 && 
              !is.na(da_part) && da_part >= 1 && da_part <= 31
  
  if (is_valid && full_yr == as.numeric(fallback_year)) {
    return(sprintf("%04d-%02d-%02d", full_yr, mo_part, da_part))
  }
  return(as.character(fallback_year))
}

# Mapping Logic
dwc_data <- tblOpnames %>%
  inner_join(tblBomenStruiken, by = "proj_cod_loc_nr") %>%
  inner_join(cdeBomenStruiken, by = c("soort" = "soort_code")) %>%
  mutate(
    # Year extraction
    extracted_year = str_extract(projectcod, "\\d{4}") %>% coalesce(
      str_extract(projectcod, "\\d{2}") %>% as.numeric() %>% 
      { ifelse(. > 50, paste0("19", .), paste0("20", .)) }
    ),
    # Full date extraction
    eventDate = map2_chr(nummer, extracted_year, extract_full_date),
    
    # Locality
    locality = paste0(coalesce(gemeente, ""), ": ", coalesce(locatie, "")) %>% str_remove("^: "),
    
    # Status Mapping
    inheems_lower = str_trim(tolower(inheems)),
    establishmentMeans = case_when(
      str_detect(inheems_lower, "[abc]") ~ "native",
      str_detect(inheems_lower, "p") ~ "introduced",
      str_detect(inheems_lower, "s") ~ "uncertain",
      TRUE ~ "uncertain"
    ),
    occurrenceRemarks = case_when(
      str_detect(inheems_lower, "a") ~ "Autochthonous origin",
      str_detect(inheems_lower, "b") ~ "Probably autochthonous origin",
      str_detect(inheems_lower, "c") ~ "Possibly autochthonous origin",
      str_detect(inheems_lower, "p") ~ "Planted / Introduced",
      str_detect(inheems_lower, "s") ~ "Spontaneous origin",
      TRUE ~ "Unknown status"
    ) %>% paste0(" (code: ", inheems, ")")
  ) %>%
  # Metadata
  mutate(
    occurrenceID = paste0("INBO:VBP:ABS:OCC:", proj_cod_loc_nr, ":", row_number()),
    basisOfRecord = "HumanObservation",
    kingdom = "Plantae",
    datasetName = "Autochtone boomsoorten en struiken in Vlaanderen",
    institutionCode = "INBO",
    collectionCode = "ABS-monitoring",
    rightsHolder = "Research Institute for Nature and Forest (INBO)",
    license = "http://creativecommons.org/publicdomain/zero/1.0/",
    language = "en"
  )
```

## 3. Coordinate Conversion

We convert Belgian Lambert 72 (EPSG:31370) coordinates to WGS84 (EPSG:4326) and round to 5 decimal places.

```{r coordinates}
# Filter records with coordinates
valid_coords <- dwc_data %>%
  filter(!is.na(coord_hor), !is.na(coord_vert))

# Convert to sf object
abs_sf <- valid_coords %>%
  st_as_sf(coords = c("coord_hor", "coord_vert"), crs = 31370) %>%
  st_transform(4326)

# Extract coordinates and round
abs_coords <- st_coordinates(abs_sf) %>%
  as_tibble() %>%
  rename(decimalLongitude = X, decimalLatitude = Y) %>%
  mutate(across(everything(), ~round(., 5)))

# Combine back
dwc_final <- valid_coords %>%
  bind_cols(abs_coords) %>%
  select(occurrenceID, basisOfRecord, scientificName = latijnse_naam, vernacularName = nederlandse_naam,
         kingdom, eventDate, year = extracted_year, decimalLatitude, decimalLongitude, 
         locality, individualCount = aantal, establishmentMeans, occurrenceRemarks,
         datasetName, institutionCode, collectionCode, rightsHolder, license, language)
```

## 4. Visualization

Interactive map showing the distribution of the first 10,000 observations.

```{r map}
leaflet(dwc_final %>% head(10000)) %>%
  addProviderTiles(providers$CartoDB.DarkMatter) %>%
  addMarkers(
    clusterOptions = markerClusterOptions(),
    popup = ~paste0("<b>", vernacularName, "</b><br>", scientificName, "<br>", occurrenceRemarks)
  )
```

## 5. Export

The final Darwin Core dataset is exported to CSV.

```{r export}
write_csv(dwc_final, "../data/dwc_occurrence_r.csv", na = "")
```

---
**Document generated by Antigravity assistant.**
